\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology

### 2017-06-09 mers-structure.tex
First draft.

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\usepackage{grffile}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

@@

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf MERS structure}

\author[1]{Gytis Dudas}
\author[2]{Luiz Max Carvalho}
\author[1]{Allison Black}
\author[1]{Trevor Bedford}
\author[2,4,5]{Andrew Rambaut}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[4]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
\affil[5]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}

@@

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus causing significant mortality and morbidity largely in the Arabian Peninsula.
The epidemiology of the virus is relatively poorly understood with most MERS patients being older males with comorbidities, although large nosocomial have occurred, notably in Jeddah, Riyadh and South Korea.
Our understanding of MERS-CoV epidemiology is further complicated by some patients not recalling contact with camels, the accepted reservoir for MERS-CoV, or any other livestock, suggesting significant cryptic MERS-CoV circulation in communities.
Seroepidemiology has been employed extensively during the time since the discovery of MERS-CoV to understand the exposure and therefore risk among camels and humans alike, but various sequencing efforts have not been fully leveraged against this zoonotic virus.
Here we use existing MERS-CoV sequencing data to estimate the number of times the virus has been introduced into humans and use the distribution of sequence clusters to estimate the reproductive number for the virus from sequence data alone.
We confirm that MERS-CoV has relatively poor capacity to spread in humans and arrive at an estimate of at least 65 zoonotic introductions of the virus into humans.

@@

\section*{Introduction}
Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
The virus, first discovered in 2012 \cite{WHO_2016}, has gone on to cause more than 1800 infections with at least 600 associated deaths.
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities.
Whilst contact with camels is often reported by patients, many do not recall contact with any livestock, suggesting a significant but unobserved community contribution to the outbreak.
Studies into MERS-CoV epidemiology often rely on serology to identify factors associated with MERS CoV exposure in potential risk groups or hosts.
The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios.
Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios, which are a fundamental part of quantifying risk.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir, nosocomial outbreak investigations and routine surveillance.

It is accepted that human MERS-CoV infections are a result of multiple introductions of the virus into humans, but no study has attempted to go beyond educated guesses about this number.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts: humans and camels.
By explicitly modelling the evolution of MERS-CoV between these two distinct host populations we show that human MERS infections are the result of frequent and unidirectional spillover events from camels into humans.
Inter-outbreak evolution of the virus occurs exclusively in camels, highlighting the need for genomic surveillance of MERS-CoV in the reservoir as well.
We also use the sequence clusters recovered from our population structure analyses to estimate the reproductive number for the virus (R$_{0}$), which is in agreement with previous findings that MERS-CoV is, on average, poor at spreading in human populations.
Our analyses also capture seasonal variation in zoonotic transmission patterns, with indications that particularly large outbreaks are the result of zoonotic transmissions that take place at the beginning of each new year.

%Here we use 307 largely complete MERS-CoV genomes, 91 of which are from camel viruses, to explicitly model the population structure of the virus in humans and camels in order to arrive at an estimate of the number of times the virus has been introduced into human populations.
%We then explore the resulting sequence clusters that can be thought of as replicates of the transmission process, albeit with different settings, as the data include large nosocomial outbreaks.

@@

\subsection*{Recombination occurs in the reservoir}
Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
Since detectable and problematic recombination in RNA viruses requires co-infection with distinct viral genotypes humans were ruled out as a suitable recombination environment \textit{a priori}.
Here we confirm this assumption by showing that evolution of MERS-CoV within human clusters is clonal.
This is not surprising, since zoonotic transmission of MERS-CoV into humans is relatively rare, of limited diversity at the point of introduction and viral diversity in humans is ultimately transient.
All of these factors act to make co-infection of humans with MERS-CoV exceedingly rare.
It is worth highlighting that MERS-CoV almost certainly recombines during human infection, but the limited diversity of virus populations in individual humans renders it invisible and of little consequence.

Although we do not detect recombination in human clusters, we see evidence for introduction of recombinant viruses into humans which evolve clonally after the transfer.
This is not unexpected either - there are strong indications that the Korean MERS outbreak was caused by a recombinant virus crossing the species barrier.
The frequency of MERS-CoV recombination in the reservoir implies that co-infection of camels is quite common, and by extension that the virus is likely to be very prevalent in this host.

\subsection*{Effects of recombination are minimal}
Most phylogenetic approaches model sequence evolution as a strictly clonal process and when recombination is present traditional phylogenetic approaches can fail to recover correct topologies, branch lengths or demographic history.
All coronavirus genera are known to be recombinogenic, and MERS-CoV is no exception, but whereas recombination is obligate for some eukaryotes, detectable RNA virus recombination can only occur during co-infection with two or more distinct viral genotypes.
Previous studies have inferred prevalent recombination in MERS-CoV and found direct evidence for it.
When not directly looking for recombination, however, most studies carried out their objectives without regard for potential effects of recombination.
Our independent analyses of the alignment split into two fragments reveals that whilst some effects of recombination are certainly detectable, they do not seem to be systematically biasing phylogenetic inference.
%So whereas we find some evidence of recombinant viruses crossing the species barrier into humans

\subsection*{Numerous cross-species transmissions of MERS-CoV}
It is generally accepted that MERS-CoV is a continuously re-emerging human pathogen, however the exact number of cross-species transmission events remains unknown.
We confirm this by identifying the camel reservoir as the host population where most of MERS-CoV evolution takes place (see figure \ref{mcc}).
The human side of MERS outbreaks, on the other hand, are caused by repetitive zoonotic transmissions from camels, which we estimated by using an explicit model of population structure.
We infer that across 216 MERS-CoV sequences collected from humans, we detect a median of 67 individual camel to human cross-species transmissions (95\% highest posterior density: 60--72), with little support for back-transmission scenarios (median 3, 95\% HPDs: 0--13).
There is also apparent heterogeneity in resulting human cluster sizes, which is not unexpected.
Large clusters of MERS are apparent and mostly correspond to known large transmission chains, often in a nosocomial setting, such as the outbreak in South Korea and Riyadh.
We also note a general predisposition for large MERS-CoV clusters to have crossed the species barrier into humans early in each year.

\begin{figure}[h]
 \centering		
	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV sequences from humans and camels.}
	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered using the structured coalescent.
	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early. 
	}
	\label{mcc}
\end{figure}

\subsection*{MERS epidemiology}
Although sequences in an outbreak necessarily represent a subset of all the cases, their clustering and resulting cluster sizes can be highly informative of underlying epidemiological processes that are otherwise inaccesible or unfeasible to study.
In our case we have used sequence data to disentangle the evolution of MERS-CoV into two discrete host compartments.
By removing the confounding effects of shared ancestry for MERS-CoV sequences we can treat every viral introduction into humans as an independent replicate of the transmission process in humans.

@@

\subsection*{MERS-CoV in camels}
Our observation of recombinant MERS coronaviruses being transmitted to humans is not cause for concern by itself, however, the mere presence of detectable recombination implies that co-infection of camels with MERS-CoV is common.
By extension, frequent co-infection implies a high prevalence of the virus in camels, which is the main unsettling conclusion.
There is ample evidence for this from seroepidemiology, which has shown that XX\% of camels are infected by age XX.
Our results show that zoonotic transmission of MERS-CoV into humans is the main drivers for the continuation of the outbreak in the Arabian Peninsula. 
Prevalence of MERS-CoV in camels could easily be translated into force of infection and given the role of zoonotic transmission as the main driver for the outbreak only reiterates the importance of controlling the MERS coronavirus in the reservoir if the human side of the outbreak is to be brought under control.

>>>>

% graphicx package, useful for including eps and pdf graphics
\usepackage{graphicx}
\usepackage{grffile}
\usepackage{subfig}
%\DeclareGraphicsExtensions{.pdf,.png,.jpg}

@@

%%% TITLE %%%
\title{\vspace{1.0cm} \LARGE \bf Structured population models demystify MERS-CoV epidemiology}

\author[1]{Gytis Dudas}
\author[2]{Luiz Max Carvalho}
\author[2,3,4]{Andrew Rambaut}
\author[1]{Trevor Bedford}

\affil[1]{Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center, Seattle, WA, USA}
\affil[2]{Institute of Evolutionary Biology, University of Edinburgh, Edinburgh, UK}
\affil[3]{Fogarty International Center, National Institutes of Health, Bethesda, MD, USA}
\affil[4]{Centre for Immunology, Infection and Evolution at the University of Edinburgh, Edinburgh, UK}

@@

Middle East Respiratory Syndrome coronavirus (MERS-CoV) is a zoonotic virus originating in camels that has been causing significant mortality and morbidity in humans in the Arabian Peninsula.
The epidemiology of the virus remains poorly understood, with hospital outbreaks, isolated cases with known exposure to camels and apparent community transmission occurring simultaneously.
Whilst traditional and seroepidemiological studies have been employed extensively throughout the epidemic, sequence data have not been utilised to their full potential in understanding transmission patterns within the outbreak.
Here we use existing MERS-CoV sequence data to explore the phylodynamics of the virus in two of its known major hosts, humans and camels.
We employ state-of-the-art structured coalescent models to show that long-term MERS-CoV evolution occurs exclusively in camels, whereas humans act as the virus' terminal host.
By analysing the distribution of human outbreak cluster sizes and zoonotic introduction times we show that MERS-CoV is unlikely to become endemic in humans and that human outbreaks in the Arabian peninsula for the most part are driven by seasonally varying zoonotic transfer of viruses from camels.

@@

\section*{Introduction}
Middle East Respiratory Syndrome Coronavirus (MERS-CoV) is a zoonotic infection of humans in the Arabian Peninsula originating from camels.
The virus, first discovered in 2012 \citep{zaki_isolation_2012,boheemen_genomic_2012}, has gone on to cause more than 1900 infections and numerous deaths, according to the World Health Organization (WHO).
Its epidemiology remains obscure, largely because outbreaks are observed among the most severely affected patients, such as older males with comorbidities \citep{assiri_2013,group_state_2013}.
Whilst contact with camels is often reported, patients may not recall contact with livestock, suggesting an unobserved community contribution to the outbreak \citep{group_state_2013}.
Studies into MERS-CoV epidemiology in the past often relied on serology to identify factors associated with MERS-CoV exposure in potential risk groups \citep{reusken_occupational_2015,reusken_2013}.
Indeed, such data have been used to show high seroprevalence in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014} and evidence of contact with MERS-CoV in workers with occupational exposure to camels \citep{reusken_occupational_2015,muller_presence_2015} compared to the general population.
Traditional epidemiological approaches have also been used to look at distributions of case clusters through time, space and across hosts \citep{cauchemez_unraveling_2016}.

Although traditional epidemiological approaches yield important clues about exposure patterns and potential for larger outbreaks, much remains opaque to such approaches.
The new era of genomic epidemiology, however, has repeatedly shown the utility of sequence data in outbreak scenarios \citep{quick_real-time_2016,arias_rapid_2016}, where data are relatively cheap to produce, can stand in for diagnostics and often yield a highly detailed picture of an epidemic when complete genome sequencing is performed consistently and appropriate metadata collected \citep{dudas_virus_2017}.
Often only sequence data can pinpoint sources of pathogens and discriminate between multiple and single source scenarios \citep{gire_genomic_2014}, which are fundamental to quantifying risk \citep{grubaugh_multiple_2017}.
Sequencing MERS-CoV has been performed as part of initial attempts to link human infections with the camel reservoir \citep{memish_human_2014}, nosocomial outbreak investigations \citep{assiri_hospital_2013} and routine surveillance \citep{park_acute_2015}.
A large portion of MERS-CoV sequences come from hospital outbreaks, where sequence data have been used to determine whether introductions were of a single origin and homogenous or a result of increased incidence \citep{fagbo_molecular_2015}.
Similar studies on MERS-CoV have taken place at broader geographic scales, such as cities \citep{cotten_2013}.

It is widely accepted that recorded human MERS-CoV infections are a result of at least several introductions of the virus into humans \citep{cotten_2013} and that contact with camels is a major risk factor for developing MERS, per WHO's guidelines.
Previous studies attempting to quantify the actual number of spillover infections have either relied on traditional, and necessarily limited, approaches \citep{cauchemez_unraveling_2016} or employed methods agnostic to signals of population structure within sequence data \citep{zhang_evolutionary_2016}.
Here we use existing MERS-CoV sequence data to investigate the population structure of the virus between two of its known hosts, humans and camels by using an explicit model of population structure \citep{vaughan_efficient_2014}.
Unlike discrete trait analyses (DTA, or ``mugration'' models) that treat a potential trait of interest as if it were a nucleotide site with an arbitrary number of states, population structure models explicitly estimate distinct branching rates in each deme separately.
By jointly estimating independent coalescence rates for MERS-CoV in humans and camels, as well as migration patterns between the two proposed demes, we show that long-term viral evolution of MERS-CoV occurs exclusively in camels.
Our results suggest that spillover events into humans are seasonal and might be associated with the calving season in camels.
Once human MERS-CoV infections are established, however, we find that MERS-CoV is poor at transmitting between humans.
Using approximate Bayesian computation we show that R$_{0}$ for MERS-CoV circulating in humans is much lower than the epidemic threshold of 1.0 and that conversely the virus has been introduced into humans hundreds of times.

@@

\subsection*{MERS-CoV is predominantly a camel virus}
The structured coalescent approach we employ \citep{vaughan_efficient_2014} identifies camels as a reservoir host where most of MERS-CoV evolution takes place (figure \ref{mcc}), whilst human MERS outbreaks are transient and terminal with respect to long-term evolution of the virus (figure \ref{exploded}).
Across 174 MERS-CoV genomes collected from humans, we detect a median of 56 individual camel to human cross-species transmissions (95\% highest posterior density (HPD): 48--63).
Whilst we identify a median of 3 (95\% HPD: 0--12) human-to-camel migrations, the 95\% HPD interval includes zero and we find that no such migrations are found in the maximum clade credibility tree (figure \ref{mcc}).
We conclude that most human-to-camel migrations are induced by the migration rate prior, whilst some are derived from phylogenetic proximity of human sequences to the apparent ``backbone'' of the phylogenetic tree.
It is most apparent in the lineages existing in early-mid 2013 that lead up to sequences comprising the MERS-CoV clade dominant in 2015, where poor sampling of MERS-CoV genetic diversity from camels allows the model to consider humans as a potential alternative host.
The first sequences of MERS-CoV from camels do not appear in our data until November 2013.
Our finding of negligible human-to-camel transmission is robust to choice of prior (figure \ref{prior}).

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (figure \ref{mcc} and \ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and robust to choice of prior.
We therefore argue that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to index cases that might be mild and ultimately not detected by surveillance or sequencing.

\begin{figure}[h]
 \centering
	\includegraphics[width=0.65\textwidth]{figures/mers_mcc.png}
	\caption{\textbf{Typed maximum clade credibility tree of MERS-CoV genomes from humans and camels.}
	Maximum clade credibility (MCC) tree showing inferred ancestral hosts for MERS-CoV recovered with the structured coalescent.
	Vast majority of MERS-CoV evolution is inferred to occur in camels (orange) with human outbreaks (blue) representing evolutionary dead-ends for the virus.
  Confidence in host assignment is depicted as a colour gradient, with increased uncertainty in host assignment (posterior probabilities close to 0.5) shown as grey.
  Some of the branches leading up to zoonotic transmission of MERS-CoV into humans are too long to have spent much of their time in humans, given our \textit{a priori} belief that MERS-CoV cannot circulate in humans for long periods of time.
  Thus branches encompassing cross-species transmissions are depicted as being in camels right up to the common ancestor or tip of each human outbreak.
	Whilst large clusters of human cases are apparent in the tree, significant contributions to human outbreaks are made by singleton sequences, likely representing recent cross-species transmissions that were caught early.
	}
	\label{mcc}
\end{figure}

\subsection*{Signals of MERS-CoV seasonality}
Unlike the most basic ``mugration'' approaches that simply infer and label the internal nodes of a phylogeny, an additional advantage structured population models have over less sophisticated approaches is the ability to recover probable distribution of migration times.
We use the posterior distribution of MERS-CoV introduction times for human clades observed in the MCC tree (figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached, leading to a sharp rise in incidence of the virus in camels that leads to seasonal variation in force of infection in humans.
Strong influx of susceptibles and subequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_seasonality.png}
	\caption{\textbf{MERS-CoV seasonality.}
A) Violin plots coloured by month showing the 95\% highest posterior density interval for odds ratios of MERS-CoV introductions into human populations.
Posterior means are indicated with circles.
Evidence for increased or decreased risk (95\% HPD excludes 1.0) for introductions are indicated by black or white circles, respectively.
Grey area spanning October to February indicates the camel calving season.
B) Each clade entering humans is positioned based on the median introduction time, as recovered by structured coalescent analyses and coloured by time of year with height indicating number of descendent sequences recovered from human cases.
95\% highest posterior density intervals for zoonotic introduction of MERS-CoV into human populations observed in the MCC tree (figure \ref{mcc}) are indicated with coloured lines, coloured by median estimated introduction time.
The black dotted line indicates the joint probability density for introductions.
Grey areas spanning October to February indicate the reported camel calving season.
	}
	\label{seasonality}
\end{figure}

\subsection*{MERS-CoV is poorly suited for human transmission}
Structured coalescent approaches clearly show humans to be a terminal host for MERS-CoV, implying poor transmissibility.
However, we wanted to translate this observation into the basic reproductive number, R$_{0}$, which is more familiar to epidemiologists, as well as extrapolate the migration process observed from sequence data to case data.
In order to achieve this we perform approximate Bayesian computation (ABC) analyses with two steps.
We simulate outbreaks using equation \ref{clusterLikelihood} with a fixed dispersion parameter and then subsample using a multivariate hypergeometric distribution from each case cluster to simulate sequencing.
Sequencing simulations take place at different levels of bias, where case clusters likely to be sequenced become even more likely to be sequenced and \textit{vice versa} for cases that would be sequenced rarely.
This is a particularly pressing issue, since \textit{a priori} we expect large hospital outbreaks of MERS to be overrepresented in sequence data, whereas sequences from index cases will be sampled exceedingly rarely.
We record the R$_{0}$ values, as well as the number of case clusters (equivalent to number of zoonotic introductions) for simulations that produce sequence cluster sizes with mean, median and standard deviation falling within the 95\% highest posterior density for these summary statistics observed in MERS-CoV data from structured coalescent analyses.

We identify R$_{0}$ values below 1.0 as being more likely to generate MERS-CoV-like distributions of sequence clusters (middle panels in figure \ref{ABC}).
For increasing levels of bias mean R$_{0}$ values observed in matching simulations are 0.844, 0.730, and 0.683, respectively.
Whilst the 95\% percentiles for R$_{0}$ values are close to 1.0 (0.720--0.995) for the unbiased sequencing simulation (\textit{i.e.} uniform sequencing efforts), we also note that increasing levels of bias are considerably better at generating MERS-CoV-like sequence clusters.
Under unbiased sequencing only 2282 (out of 363 000) simulations fit our observations in structured coalescent analyses of MERS-CoV which increase to 9763 and 9923 for bias levels of 2.0 and 3.0, respectively.
This is apparent in the heights of the histograms in figure \ref{ABC}.
Simulations with bias suggest that R$_{0}$ values could have ranges [0.605--0.870] (bias = 2) and [0.555--0.825] (bias = 3).

Lower values for R$_{0}$ in turn suggest relatively large numbers of zoonotic transfers of viruses into humans.
The median number of cross-species introductions observed in simulations matching MERS data without bias are 339 (95\% percentiles 255--431).
These numbers jump up to 558 (95\% percentiles 424--720) for bias = 2 and 650 (95\% percentiles 481--850) for bias = 3 simulations, which as mentioned previously seem to produce considerably more MERS-CoV-like data.
Our results suggest a large number of unobserved MERS index cases.
Given this, we also performed simulations where the total number of cases is doubled to 4000 to explore the impact of dramatic underestimation of MERS cases.
In those sets of analyses R$_{0}$ values tend to decrease even further as numbers of introductions go up, although very few simulations match currently observed MERS-CoV sequence data (figure \ref{extra_cases}).
Overall our analyses indicate that MERS-CoV is poorly suited for human-to-human transmission, with an estimated R$_{0}<1.0$ and sequence sampling wholly unrepresentative of viral diversity found in humans.
Given these findings, and the fact that large outbreaks of MERS occurred in hospitals, the combination of frequent spillover of MERS-CoV into humans and occasional outbreak amplification via poor hygiene practices appear sufficient to explain observed epidemiological patterns of MERS-CoV.

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_epi.png}
	\caption{\textbf{Results of ABC simulations.}
Each row corresponds to a different bias value used to concentrate the hypergeometric distribution that simulates sequencing, and goes from 1 (no bias) to 2, and 3 (increasing levels of bias).
Leftmost scatter plots show a 10\% subsample of all outbreak and sequencing simulations, coloured by the R$_{0}$ value used for the simulation.
The dotted rectangle identifies the 95\% highest posterior density bounds for sequence cluster size mean and standard deviation observed for MERS-CoV data.
The distribution of R$_{0}$ values found within the dotted rectangle is shown in the middle, on the same y axis across all levels of bias.
Bins falling inside the 95\% percentiles are coloured by R$_{0}$, as in the leftmost scatter plot.
The distribution of total number of introductions associated with simulations matching MERS-CoV sequence clusters is shown in the plots on the right, on the same y axis across all levels of bias.
Darker shade of grey indicates bins falling within the 95\% percentiles.
ABC indicates R$_{0}$ for MERS-CoV is likely to be below 1.0, with biased sequencing and numbers of zoonotic transmissions numbering in the hundreds.
	}
	\label{ABC}
\end{figure}

%\begin{figure}[h]
% \centering
%	\includegraphics[width=0.65\textwidth]{figures/}
%	\caption{\textbf{}
%
%	}
%	\label{}
%\end{figure}

\subsection*{Recombination shapes MERS-CoV diversity}
Recombination has been shown to occur in all genera of coronaviruses, including MERS-CoV.
In order to explore the role of recombination in shaping MERS-CoV genetic diversity we used two recombination detection approaches across partitions of data corresponding to inferred MERS-CoV clades.
Both methods rely on sampling parental and recombinant alleles within the alignment, although each quantifies different signals of recombination.
One hallmark of recombination is the ability to carry alleles derived via mutation from one lineage to another, that appear as repeated mutations taking place in the recipient lineage, somewhere else in the tree.
PHI (pairwise homoplasy index) test quantifies the appearance of these excessive repeat mutations (homoplasies) within an alignment \citep{bruen_simple_2006}.
Another hallmark of recombination is spatial clustering of alleles in the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
3Seq relies on the spatial structure of nucleotide similarities between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences \citep{boni_exact_2007}.

Both tests can give spurious results in cases of extreme rate heterogeneity and sampling over time \citep{dudas_mers-cov_2016}, but both tests have not been reported to fail simultaneously.
PHI and 3Seq methods consistently identify most of the apparent `backbone' of the MERS-CoV phylogeny as encompassing sequences with evidence of recombination (figure \ref{recombination_tree}).
Neither method can identify where in the tree recombination occurred, but each full asterisk in figure \ref{recombination_tree} should be interpreted as the minimum partition of data that still captures both donor and recipient alleles involved in a recombination event.
This suggests a non-negligible contribution of recombination in shaping existing MERS-CoV diversity.
As before \citep{dudas_mers-cov_2016}, we show large numbers of homoplasies in MERS-CoV data (figure \ref{incompatibilities}) with some evidence of spatial clustering of such alleles.
Homoplasies present in multiple strains at a time, indicating ancestral events that have been successful.
Though the proportion of time that the virus is inferred to spend in camels alone may be sufficient to argue that camels are the host where MERS-CoV recombines, incidence of MERS-CoV is known to be much higher in camels \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2014,ali_systematic_2017}.
This provides ideal conditions for co-infection, which is a pre-requisite for RNA virus recombination to occur.

Conversely, our results argue strongly that co-infection of humans with distinct lineages of MERS-CoV should be exceedingly rare.
We find little evidence that recombination might be interfering with human outbreak cluster inference beyond what could be expected of phylogenetic noise (figure \ref{recombination}A).
And whilst we observe evidence of distinct phylogenetic trees from different parts of the MERS-CoV genome (figure \ref{recombination}B), human clades are minimally affected and large portions of the posterior probability density in both parts of the genome are concentrated in shared clades (figure \ref{flower}).
Observed departures from strictly clonal evolution suggest that whilst recombination is a serious issue in MERS-CoV phylogenies, its effects on the human side of MERS outbreaks is minimal, as expected.
Overall we advise extreme caution when interpreting branches in MERS-CoV phylogenies that are suspected of having spent any time in camels, especially when attempting to identify origins of human cases.
Conversely, the largely clonal evolution of MERS-CoV in humans, despite limited diversity, should be more amenable to traditional phylogenetic approaches.

% \begin{figure}[h]
%  \centering
% 	\includegraphics[width=0.65\textwidth]{figures/mers_chain.png}
%   \includegraphics[width=0.65\textwidth]{figures/mers_fragments.png}
% 	\caption{\textbf{Tangled chain of MERS-CoV genome and its parts.}
% 	Tangled chain of the full genome (a), all nucleotides up to position 21000 (b) and all nucleotides past position 21000 (c) of MERS-CoV.
% 	The same taxa, if available in neighbouring trees, are connected via coloured lines (human sequences in blue, camel sequences in red).
% 	Tree branches are coloured by inferred ancestral host state (human in grey, camel in black).
% 	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in the camel host.
% 	Human clusters in grey change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
% 	This is evidence for recombinant viruses generated in the reservoir entering human populations.
% 	Coding sequences in the MERS-CoV genome are indicated with arrows at the bottom of the plot with position 21000 indicated by the black line.}
% 	\label{mers_chain}
% \end{figure}

% \begin{figure}[h]
% \centering
%
% 	\caption{\textbf{Pairwise co-occurence matrix for human clusters.}
% Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
% Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
% Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
% 	}
% 	\label{coocurrence}
% \end{figure}
\begin{figure}%
    \centering
    \subfloat{{\includegraphics[width=7cm]{figures/mers_fragments.png} }}%
    \qquad
    \subfloat{{\includegraphics[width=5cm]{figures/mers_chain.png} }}%
    \caption{\textbf{Recombinant features of MERS-CoV phylogenies}.
    A) Heatmap shows the posterior probability that a pair of tips from trees of genomic fragments fall within the same clade - tips from fragment 1 are on the x axis, tips from fragment 2 are on the y axis.
    Tips are ordered by their appearance in tree of genome fragment 2 (positions from nucleotide 21000 onwards) reduced to just the human tips and coloured by inferred host (blue for human, orange for camel) on the left.
    Human clusters are largely well-supported as monophyletic and consistent across trees of both genomic fragments.
    B) Phylogenies derived from MERS-CoV genome nucleotides up to position 21000 (left) and all nucleotides past position 21000 (right), reduced to just the human tips.
  	Same taxa are connected via coloured lines indicating their vertical position in the tree on the right.
  	Tree branches are coloured by inferred ancestral host state (human in blue, camel in orange).
  	Whilst some of apparent incongruities are caused by having less data in some of the fragments, inconsistencies between topologies occur across internal branches inferred to be in camels.
  	Human clusters in blue change phylogenetic positions between the trees wholesale, with minor incongruences within clusters.
  	This is evidence for recombinant viruses generated in the reservoir entering human populations.
    }%
    \label{recombination}%
\end{figure}

\subsection*{Demographic trajectory of MERS-CoV in camels}
Since MERS-CoV in humans evolves under a distinct demographic regime, we split our sequence dataset based on whether sequences were collected in humans or camels.
We supplement camel sequence data with a single earliest sequence from each human cluster, treating viral diversity present in humans as a sentinel sample of MERS-CoV diversity circulating in camels.
This removes conflicting demographic signals sampled during human outbreaks, where densely sampled closely related sequences from humans could be misconstrued as evidence of demographic crash in the viral population.

Our attempt at controlling the confounding effects of different sampling regimes between humans and camels revealed a troubling pattern, however.
We find that this camel-like dataset results in poor MCMC performance, where independent chains reach stationary distributions but do not necessarily converge on the same distributions.
Of the five runs we set up two chains converged to one stationary distribution, two to another and the last one to a third.
Demographic trajectories recovered by the two main stationary distributions are very similar and differences between the two appear to be caused by convergence onto different tree topologies, with onward effects on evolutionary rate estimates and distributions of coalescence times.
This non-convergence effect may have been masked previously by the use of all available MERS-CoV sequences from humans which may have nudged MCMC towards one of the multiple stationary distributions.

Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in MERS-CoV relative genetic diversity (N$_{e}\tau$) over time (figure \ref{skygrid}).
However, we do note that estimates of relative genetic diversity are relatively low overall, and MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
This raises the possibility that MERS-CoV populations undergo lineage turnover at a rate not often seen in other livestock viruses (\textit{e.g.} swine influenza A viruses \citep{vijaykrishna_long-term_2011}).
Part of this could be caused by MERS-CoV ability to re-infect camels previously exposed to the virus \citep{ali_systematic_2017}, which may result in a process akin to antigenic drift, where the depletion of susceptible calves leads to ever increasing selective pressure for the virus to evolve antigenically.
Whilst this could be tested by looking for seasonally varying patterns in selection on MERS-CoV surface proteins, recombination would remain a key confounding problem.

% In order to disentangle distinct demographic modes within the phylogeny of MERS-CoV we split our alignment according to whether the virus was collected from camels or humans.
% For sequences collected from humans and inferred to be in the same cluster we only keep the earliest sequence, and arrive at a sequence dataset that for the most part is comprised of viral diversity present in camels, and supplemented with human sequences that are treated as sentinel samples.

@@

\subsection*{Role of population structure in phylodynamics}
Population structure yields important insight into any biological system.
Population structure often implies the existence of barriers experienced by an organism, be it geographic \citep{dudas_virus_2017}, ecological \citep{smith_dating_2009} or evolutionary \citep{turner_genomic_2005,dudas_reassortment_2015}.
Understanding what barriers exist and how they affect transmission of viruses remains one of the key parameters in designing efficient interventions.
In this study we aimed to understand the drivers of MERS coronavirus transmission in humans and what role the camel reservoir plays in perpetuating the epidemic in the Arabian peninsula.
Whilst there are clear indications that human cases of MERS are largely the result of sporadic spillover from camels, sequence data have not yet been fully utilised in estimating basic epidemiological parameters, such as R$_{0}$ or the number of zoonotic introductions of MERS-CoV into humans.

We showed that currently existing models of population structure \citep{vaughan_efficient_2014} can identify distinct demographic modes in MERS-CoV data, where viruses continuously circulating in camels repeatedly jump into humans and cause small outbreaks doomed to extinction (figures \ref{mcc}, \ref{exploded}).
This inference succeeds under different choices of priors for unknown demographic parameters (figure \ref{prior}) and in the presence of strong biases in sequence sampling schemes (figure \ref{ABC}).
From sequence data we identify at least 50 zoonotic introductions of MERS-CoV into humans from the reservoir (figure \ref{mcc}), from which we extrapolate that hundreds more such introductions must have taken place (figure \ref{ABC}).
We also looked at potential seasonality in MERS-CoV spillover into humans, which is an additional advantage of explicitly modeling population structure.
Our analyses indicated a period of three months where the odds of a sequenced spillover event are increased, with timing consistent with an enzootic amongst camel calves (figure \ref{seasonality}).
As a result of our identification of large and asymmetric flow of viral lineages into humans we also find that the basic reproduction number for MERS-CoV in humans is well below the epidemic threshold (figure \ref{ABC}).
Poor capacity of MERS-CoV to spread human-to-human explains why evidence of substantial community transmission has been lacking, as well as the absence of virus circulation outside of Arabian peninsula.
Perhaps a clearer indicator of this has been the lack of a global MERS pandemic, which would have certainly taken place had R$_{0}$ been above the epidemic threshold.
